 

### 2021-04-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al. 2014*].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

@@

The amount of light reaching lower canopy, understories and forest floors are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks. Shaded canopy layers depend on sunflecks, where they receive 10-80% of their photon flux density for photosynthesis in short bursts within minutes of light exposure [@waySunflecksTreesForests2012]. Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

These light regimes shape foliage at a vertical gradient, which in turn shapes the microenvironment. At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and the overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming  [@fauset_differences_2018; @millerOnlySunlitLeaves2021; @michaletzPlantThermoregulationEnergetics2015; *Defrenne et al. 2021*]

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom. Understory wind-speed is also substantially reduced with increased canopy height [@juckerCanopyStructureTopography2018]
Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].

@@

The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al. 2021, @davis_microclimatic_2019, @zellweger_seasonal_2019]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with canopy cover [S. R. Hardwick et al. 2015; @juckerCanopyStructureTopography2018]. Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999]. During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to radiative heat loss from the soil caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature. However, beyond a certain height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018]. At a species level, broadleaf trees with higher shade casting ability also contribute to greater temperature buffering [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a].

Seasonally, the capacity to moderate buffering is greater with water availability. During the wet season, tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al. 2017], have all shown to offset temperatures positively in the understory, with higher maximum air temperatures and relative humidity above forest canopies and gaps than in below canopies and understories. During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016]. Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

@@

Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al. 2021], however, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed. [@bertrandCommentForestMicroclimate2020]

@@

With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions. In European forests, buffering capacity is stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a]. Trees in European high altitude forests have shown to maintain sparser stands [@vonarxSpatiotemporalEffectsForest2012a] and open canopies with greater light interception through canopy layers, and to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015]. While increase in tree height (in closed-forests) increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate humid-mountain forests in China, tree height decreased with 
altitude, whereas tree growth sensitivity (based on tree-rings) to climate change increased [@liangForestTypeHeight2019]. Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a, @fauset_differences_2018].The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019]. Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davis_microclimatic_2019; @macekMaximumAirTemperature2019]

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020] 
The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of radiation input versus heat lost to the environment [Fig. `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
In open-canopy needle-leaf forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially higher than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021, @martin_boundary_1999, @zweifelMiddayStomatalClosure2002]  
Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
For instance, during the dry season in a semi-decidudous tropical moist forest in Panama [@rey-sanchez_spatial_2016]. Sun leaves can also be cooled relative to $T_{air}$ more than shade leaves [@rey-sanchez_spatial_2016]. 
At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height. Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
- In a tropical wet and moist forests of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a] 
-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al. 2019)

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al. 2014*].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities].
Individually or in combination, these changes have the potential to reduce forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020]. 
The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

@@

The amount of light reaching lower canopy, understories and forest floors are highly variable, ranging from high light gaps to deeply shaded understory regions [@chazdonPhotosyntheticLightEnvironments1984; @tymenQuantifyingMicroenvironmentalVariation2017; @smithSeasonalDroughtrelatedChanges2019]. Forest floors of tropical forests with densely packed overstory trees, receive approximately 1-2% of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]and needle-leaf forest floors receive 17-34% of incoming sunlight [@baldocchiSeasonalVariationEnergy1997]. Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute to the light environment of shaded forest regions, in the form of sunflecks. Shaded canopy layers rely on sunflecks for photosynthesis, where they receive 10-80% of their photon flux density in short bursts within minutes of light exposure [@waySunflecksTreesForests2012]. Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

These light regimes shape foliage at a vertical gradient, which in turn shapes the microenvironment. At a foliar level, resulting light-adaptation in the form of sun and shade leaf dichotomy also facilitates plant-adaptation to the light environment in the understory and the overstory, ultimately contributing to shaping the microenvironment, forest vertical-profile's thermal sensitivity and their response to warming  [@fauset_differences_2018; @millerOnlySunlitLeaves2021; @michaletzPlantThermoregulationEnergetics2015; *De Frenne et al. 2021*]

### Wind speeds 
**Wind speeds are also higher at the top of the canopy (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`), owing to... [@juckerCanopyStructureTopography2018].**
Across all **#** forested NEON sites, wind speeds increased with height, with daily maximum wind speeds at the top of the averaging up to **#** m s^-1^ higher at the top of the vertical profile than at the bottom. Understory wind-speed is also substantially reduced with increased canopy height [@juckerCanopyStructureTopography2018].
Consistent with these results from NEON sites, higher wind speeds at greater heights within a forest canopy have been observed in both closed canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] and open savannas [@curtis_intracanopy_2019].

@@

The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al. 2021; @davis_microclimatic_2019; @zellweger_seasonal_2019]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with canopy cover [S. R. Hardwick et al. 2015; @juckerCanopyStructureTopography2018]. Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies. They maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999]. During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to radiative heat loss from the soil caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature. However, beyond a certain height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018]. At a species level, broadleaf trees with higher shade casting ability also contribute to greater temperature buffering [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a].

Seasonally, the capacity to moderate buffering is greater with water availability. During the wet season, tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al. 2017], have all shown to offset temperatures positively in the understory, with lower maximum air temperatures and relative humidity below canopies and understories than above forest canopies and gaps. During the dry season of a semi-deciduous tropical forest in Panama, however, maximum temperatures were similar in the upper and lower canopy, partially because in the dry season some canopy trees are leafless, resulting in more light reaching the sub-canopy, and more boundary layer turbulence [@rey-sanchez_spatial_2016]. Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

@@

Thus, we see that buffering by canopy cover smooths the affect temperature extremes and fluctuation in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020; @rollinsonClimateSensitivityUnderstory; Defrenne et al. 2021],though, as a result of global rise in temperatures, inevitable increases in understory temperatures continue to be observed. [@bertrandCommentForestMicroclimate2020]

@@

With higher altitude, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions. In European forests, buffering capacity is stronger in low altitude forests than high altitude forests [@vonarxSpatiotemporalEffectsForest2012a]. Trees in European high altitude forests have shown to maintain sparser stands [@vonarxSpatiotemporalEffectsForest2012a] and open canopies with greater light interception through canopy layers, and to the understory [@rajsnerovaMorphologicalBiochemicalPhysiological2015]. While increase in tree height (in closed-forests) increases buffering in the understory [@juckerCanopyStructureTopography2018], in timberline and temperate humid-mountain forests in China, tree height decreased with 
altitude, whereas tree growth sensitivity (based on tree-rings) to climate change increased [@liangForestTypeHeight2019]. Less capacity to buffer climatic conditions in open, high altitude forests might thus have implications to forest regeneration [@vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018].The strength of buffering also varies with respect to distance to the coast, topographic position, elevation [@zellweger_seasonal_2019]. Riparian forests tend to sustain buffering, protecting understory temperatures even during increasing temperature extremes compared to open areas by a stream [@davis_microclimatic_2019; @macekMaximumAirTemperature2019]

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020] 
The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of energy as radiation input versus heat lost to the environment [Fig. `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
In open-canopy needle-leaf forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially warmer than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]  
In closed-forests, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
For instance, during the dry season in a semi-decidudous tropical moist forest in Panama [@rey-sanchez_spatial_2016]. Sun leaves can also be cooled relative to $T_{air}$ more than shade leaves [@rey-sanchez_spatial_2016]. 
At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height. Additionally, with increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
- In a tropical wet and moist forests of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a] 
-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al. 2019)
